---
title: "Network Inference Prep"
author: "Kevin Hu"
date: "February 12, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Introduction

At this point in time, we have a valid and successful simulation for de novo exomic mutation in both cases and controls for a population for an arbitrarily designated disease. We also have a working implementation of PINTS. We've investigated a general model of heterogeneous disease, to little success. At this point in time, we're going to move on to network inference again, and see if PINTS is able to detect signal nodes that would not be detected directly. 

This file is strictly for the collection and commenting of the code that'll be used to address the above problem. This includes collecting all the code for simulating individuals, running PINTS, loading the relevant data, and simulating heterogeneous disease. Here we go. 

### Packages

```{r}
install.packages("devtools")
library(devtools)

install.packages("RSQLite")
install.packages("XML")

install_github("CotsapasLab/PINTSv1")
library(PINTS)

#Installing from CRAN
install.packages("igraph")
library(igraph)

#installing from Bioconductor.org
source("http://bioconductor.org/biocLite.R")
biocLite("BioNet")
biocLite("RBGL")
biocLite("graph")
biocLite("Matching")
library(BioNet)

library(HighSpeedStats)
```

### PINTS

We start with the initial PPI network from the InWeb3 dataset ([K. Lage et al][1]), which is included in the package. This contains a *pajek* file graph file and a network node annotation file with Ensemble and HUGO annotations, the former of which is used as the node names. loadPPINetwork() loads the two above into a list consisting of a graph object (igraph) and data frame of node annotations. 


```{r}
PPINetPath <- system.file("extdata", "InWeb3_HC_PPI_comp0.net", package="PINTS")
PPIAnnotPath <- system.file("extdata", "InWeb3_HC_PPI_comp0_geneAnnotation.txt", package="PINTS")
PPI <- loadPPINetwork(PPINetPath, PPIAnnotPath)
network <- PPI$g #our network file
nodeAnnot <- PPI$nodeAnnot #data frame of node annotations
```

We then apply node weights to the network - original PINTS uses mutational constraint scores, but we're going to be using the mutation count from each simulation, or perhaps the p-value for each gene from those simulations. 

```{r}
#First get the requisite genes in there.
geneScorePath <- system.file("extdata", "InWeb3_HC_PPI_Evolution_pvals.txt", package="PINTS")
nodeWeightTable <- read.table(geneScorePath, head=T, stringsAsFactors=F)
#gene expression data
geneExpressionPath <- system.file("extdata", "Binary_Pref_Gene_Exp_Roadmap.txt", package="PINTS")
gExpData <- read.table(geneExpressionPath, head=T, stringsAsFactors=F)
#And this tells us which genes are actually expressed in the tissues of interest. 
genes <- gExpData[,"Ensemble"] 
```

This portion simulates a random walk on the PPI network, and thus can choose for us a subnetwork for a candidate heterogeneous disease. Parameters can be adjusted to get networks of different spreads and sizes. This will probably be done at the beginning of a simulation to choose the genes to be simulating. 

```{r}
######################################################################
#IMPORTANT PARAMETERS
subnet.type <- "walk" #can choose walk or neighborhood
#ncp_mean <- mean
#ncp_Var <- var
nSubnet <- 15
#IMPORTANT POUND SIGNS
######################################################################

#We pick a node at random and perform a random walk of length nSteps. The function random_walk() has an argument stuck which is set to "return", which returns the partial walk in the case that it gets stuck. This is still useful as it gives information about structure. 

#Number of genes in the network
nGenes <- length(nodeAnnot[[3]])
startNode <- sample(1:nGenes, 1)
startGenes <- nodeAnnot[startNode, c('Ensemble')]
#Creates random subnet
if(subnet.type == "walk"){
  subnet <- random_walk(network, start = startGenes, steps = nSubnet)
}

if(subnet.type == "neighborhood"){
  if(neighborhood.size(network, 1, nodes = startGenes, mindist = 0) <= nSubnet){
    subnet <- neighborhood(network, 1, nodes = startGenes, mindist = 0)
  } else{
    subnet <- neighborhood(network, 0, nodes = startGenes, mindist = 0)
    k <- 0
    while(k < nSubnet){
      subnet[[1]] <- c(subnet[[1]], neighborhood(network, 1, nodes = startGenes, mindist = 0)[[1]][k])
      k <- k+1
    }
  }
}
#subnet <- neighborhood(network, 1, nodes = startGenes, mindist = 0)
print(subnet)
```

This portion assigns weights to all the nodes depending on if they're signal or background. We'll change this portion of the code so that we pull in p-values from the simulation rather than doing this random thing. I need to figure out what exactly I was doing first, though. 

```{r}
#Then we assign our random weights to all nodes
for(i in 1:length(nodeAnnot[["Ensemble"]])) {
  if(sample(1:2, 1) %% 2 == 0){
  nodeWeightTable[i, c('pvalues')] <- 0.05
  }
  else{
    nodeWeightTable[i, c('pvalues')] <- 0.05
  }
}
#and significant weights for signal nodes
for(i in 1:nSubnet) {
  nodeWeightTable[subnet[[i]], 2] <- 5e-12
  print(nodeWeightTable[subnet[[i]],2])
}

#Our simulated disease network with prepicked subnet. 
diseaseNetwork <- buildDiseaseNetwork(network, nodeWeightTable, nodeAnnot, nodeNameBy ="Ensemble", nodeSelected=genes)
```

This portion then does the actual subnetwork search. It creates null disease networks, compares the one we have against those, and then computes a p-value for the significance of our candidate network. I think that's how it works. I need to look at the default implementation to see what exactly is going on. 

```{r}
#Create a set of null disease networks to determine if our simulation has power. 
set.seed(1234)
nNull <- 100 #the number of null top subnetwork  
nullDiseaseNetwork <- randDiseaseNetwork(diseaseNetwork, nPermute=nNull)
print(nullDiseaseNetwork[1:10,1:12])

result <- runDiseaseSubnetSearch(diseaseNetwork, nodeAnnot, sigThres= 5e-6, nullDiseaseNetwork, showResult=TRUE)
subnetNodes <- result[[1]]

subnetNodesHUGO <- nodeAnnot[ which(nodeAnnot[,"Ensemble"] %in% subnetNodes), "HUGO"]
topSubnet <- induced.subgraph(diseaseNetwork, subnetNodes)
is.connected(topSubnet)
topSubnet.layout <- layout.fruchterman.reingold(topSubnet)
plot(topSubnet, layout = topSubnet.layout, main= "The MC-associated subnetwork", vertex.color="skyblue", vertex.size= 10, vertex.label=subnetNodesHUGO, vertex.label.font=3, vertex.label.cex=0.5, edge.width=2)
```

This is an implementation of checking whether disease nodes were detected by PINTS or not. It doesn't work very well. We will probably chuck it in lieu of some better checking method. Not sure what that will be, but it will work, is what's important. 

```{r}
count <- 0
for(i in 1:nSubnet){
  for(j in 1:length(result[[1]])){
    if (grepl(nodeAnnot[as.character(subnet[[i]]),3],result[[1]][j])){
      print(nodeAnnot[as.character(subnet[[i]]),4])
      count <- count + 1}
  }
}
if(count == 0){
  print("WHY IS THIS NOT WORKING")
}else{
  print(count)
}

find_gene <- function(table, name){
  for(i in 1:length(table[[2]])){
    if (grepl(nodeAnnot[i,3],name)){
      print(nodeAnnot[i,])
      return(nodeAnnot[i,1])
    }
  }
}

```


### Cohort Simulation

The required data files are:

* mut_data.txt
* control_summary_10k.txt, or another source of control mutation counts.
* genes_of_interest.txt, or another source of possible disease genes. 

mut_data.txt is modified in the code to provide the list of genes common to both Ensembl and InWeb. We then pull out the probability of nonsynonymous mutation. 

```{r}
gene_probs <-
  read.table(
    "mut_data.txt",
    header = TRUE,
    stringsAsFactors = FALSE,
    sep = ","
  )
gene_list <-
  data.frame(
    Ensembl = gene_probs$Ensembl,
    freq = gene_probs$mu_mis + gene_probs$mu_lof,
    stringsAsFactors = FALSE
  )
rm(gene_probs)

control_agg <- read.table("control_summary_10k.txt", sep = "\t", header = TRUE)

gois <- read.table("genes_of_interest.txt", sep = "\t", header = TRUE, as.is = TRUE)
```

Below is all the code used in the simulation. Included are some of the functions used in doing direct testing of each individual gene, which will not be used. However, it's included for the sake of completion, and so I would actually finish annotating everything. 

```{r}
# Generate a population with case_number # of cases and control_number # of controls, at a base non-synonymous de novo mutation rate of freq. Returns a nx3 data frame.
Gen_pop <- function(case_number, control_number, freq) {
  ID <- seq(1, case_number + control_number, 1)
  case_mutations <- rpois(case_number, freq)
  control_mutations <- rpois(control_number, freq)
  
  for (i in 1:case_number) {
    while (case_mutations[i] == 0) {
      case_mutations[i] <- rpois(1, freq)
    }
  }
  
  cc <- c(rep(TRUE, case_number), rep(FALSE, control_number))
  
  cohort <- data.frame(ID, cc, c(case_mutations, control_mutations))
  colnames(cohort) <- c("ID", "cc", "mutations")
  
  return(cohort)
}

# Isolates a new data frame from the population data frame that has just the individual ID and the case/control status. Returns an nx2 data frame.
Gen_pop_info <- function(population) {
  pop_info <- data.frame(population$ID, population$cc)
  
  return(pop_info)
}

# Extends the population data frame to have columns for each possible gene that can receive a mutation. Returns an nx(3+g) data frame. 
Initiate_population <- function(population, gene_list) {
  indivs <- rep(0, length(population[[1]]))
  result <- cbind(replicate(length(gene_list), indivs))
  
  result1 <- data.frame(population, result)
  
  names <- c("ID", "case_control", "num_mutations", gene_list)
  colnames(result1) <- names
  
  return(result1)
}

# Given a row from the population, it randomly distributes the mutations amongst the space of genes conditional on case/control status. Returns a row vector.
Dist_mut <-
  function(indiv,
           gene_list,
           gene_freq,
           gene_case_list,
           gene_case_freq) {
    if (indiv[2] == TRUE) {
      mutation <- which(rmultinom(1, 1, gene_case_freq) == 1)
      indiv[as.character(gene_case_list[mutation])] <- 1
      if (indiv[3] > 1) {
        mutation <- rmultinom(1, as.integer(indiv[3]) - 1, gene_freq)
        for (i in which(mutation > 0)) {
          indiv[gene_list[i]] <- mutation[i]
        }
      }
      return(indiv)
      
    }
    else{
      if (indiv[3] == 0) {
        return(indiv)
      }
      else{
        mutation <- rmultinom(1, as.integer(indiv[3]), gene_freq)
        for (i in which(mutation > 0)) {
          indiv[gene_list[i]] <- mutation[i]
        }
        return(indiv)
      }
    }
    
  }

# Function to take in a population data frame and output the same data frame with mutation data. Iterates Dist_mut through apply(). Returns an nx(g+3) data frame. 
Cohort_mut <-
  function(population,
           gene_list,
           gene_freq,
           gene_case_list,
           gene_case_freq) {
    population <- t(apply(population,
                     1,
                     Dist_mut,
                     gene_list,
                     gene_freq,
                     gene_case_list,
                     gene_case_freq))
    #new_pop <- t(new_pop)
    return(population)
  }

# Calculates the probability that a gene is mutated more in cases than controls, under the null that there is no difference between the two. Returns numeric.
Exact_test <- function(gene_mut, case_control) {
  elbat <- table(gene_mut, case_control)
  return(feTestR(elbat[4], elbat[2], elbat[3], elbat[1]))
}

# Function that iterates Exact_test over all the genes. Returns nx2 data frame of p-values. 
Test <- function(population, population_info, num_perms) {
  print("Starting:")
  proc.time()
  gene_id <- colnames(population)[4:length(population[1, ])]
  num <- length(gene_id)
  peoples <- population
  observed <- rep(0, length(gene_id))
  
  permat <- matrix(1, nrow = num, ncol = num_perms + 1)
  
  condition <- as.vector(colSums(population) > 0)
  print("Everything initialized")
  proc.time()
  
  #Make initial observed p-values
  for (i in 1:num) {
    if (condition[i + 3]) {
      permat[i, 1] <-
        Exact_test(peoples[, i + 3], peoples[, 2])
    }
  }
  
  #iterate
  for (i in 1:10) {
    peoples[, 2] <- sample(population_info[, 2])
    for (j in 1:num) {
      if (condition[j + 3]) {
        permat[j, i + 1] <-
          Exact_test(peoples[, j + 3], peoples[, 2])
      }
    }
  }
  
  p.val <- rep(1, num)
  for (i in 1:num) {
    p.val[i] <- sum(permat[i, -1] <= permat[i, 1]) / num_perms
  }
  
  print("final p-vals completed:")
  proc.time()
  
  return(data.frame(gene_id, p.val))
}

# wrapper runs simulations of populations with certain permutations of disease genes against a fixed control background. This is just used for the direct significance calculation. Returns a giant data frame with information on each specific gene, its heterogeneity, the case size, and p-value. We won't really use this. 
#gois is the list of genes of interest with the pLI
#Het is a matrix where each row is the relative proportion of cases it explains, so the length of the row is the number of disease genes there are.
#n is case sample size
#i is number of iterations per parameter set
#controls is aggregate of controls being used
wrapper <- function(gois, Het, n, i, controls) {
  gene_num <- length(Het[1,]) #number of genes that are causal
  long <- t(permutations(length(gois[, 1]), gene_num)) #set of permutations
  
  gene <- rep(as.vector(long), i * length(Het[,1]))
  
  het <- c()
  for (j in 1:length(Het[, 1])) {
    het <- c(het, rep(Het[j,],length(long[1,])))
  }
  het <- rep(het, i)

  num <- n
  
  results <-
  data.frame(gene, het, num, pvals = NA)
  #head(results)
  
  for (j in 1:(length(long[1,]) * length(Het[,1]) * i)) {
    population <- Gen_pop(n, 1, 0.38)[1:n,]
    population <-
      Initiate_population(population, gene_list$Ensembl)
    population <-
      Cohort_mut(population,
                 gene_list$Ensembl,
                 gene_list$freq,
                 gois[results$gene[(gene_num * (j-1) + 1):(gene_num * j)], 1],
                 results$het[(gene_num * (j-1) + 1):(gene_num * j)])
    cases <- colSums(population)
    
    for(q in 1:gene_num){
    results$pvals[gene_num * (j-1) + q] <-
      big_fish(as.character(gois[results$gene[gene_num * (j-1) + q], 1]), cases, controls)
    }
    if(j %% 1000 == 0){
      print(2*j)
    }
  }
  
  return(results)
}

# I'm not sure how this is different from Exact_test, to be honest. Gotta look into this later. 
big_fish <- function(gene, cases, controls) {
  ill_case <- as.integer(cases[gene])
  well_case <- as.integer(cases[2] - cases[gene])
  ill_control <- as.integer(controls[controls[, 1] == gene, 2])
  well_control <- 500000 - ill_control
  
  x <- feTestR(ill_case, well_case, ill_control, well_control)
  return(x)
}

# Generates a heterogeneity matrix for h genes with n different iterations. 
gen_het <- function(h, n){
  dirtframe <- matrix(rep(0, n * h),nrow = n)
  final <- NA
  for(i in 1:n){
    short <- runif(h-1, 0, 100)
    long <- c(short, 0, 100)
    ordered <- sort(long)
    for(j in 1:h){
      final[j] <- ordered[j+1] - ordered[j]
    }
    dirtframe[i,] <- final
  }
  
  return(dirtframe)
}
```


